### 2013-08-01 Tex/chapters/methods.tex
Fehler und inkonsistenzen behoben
Überblick über die Observablen gegeben

\subsection{Thermodynamic Theory}
\label{ssec:theory}
    The disordered Ising model will be examined as a \emph{canonical system} in
    \emph{equilibrium}. A canonical system can exchange energy with a
    heat bath, thus has a constant temperature -- the temperature of the
    heat bath.
    Equilibrium is defined as a steady state, where
    the observables are only fluctuating but not changing in any
    particular direction. For example thermal equilibrium denotes the
    condition, that the system under scrutiny has reached the temperature
    of the heat bath. This is the case, once no energy is exchanged
    between them, thus the energy of the simulated system reaches a
    steady state.
    In a canonical ensemble the probability \(p_i\) of a state
    \(i\) is distributed according to a Boltzman distribution
    \begin{equation}
        p_i \propto e^{-\beta H_i}
    \end{equation}
    \begin{equation}
        \beta = \frac{1}{k_B T}
    \end{equation}
    Further the free energy \(F\) of a canonical ensemble is minimized
    in equilibrium and all observables can be derived from \(F\)
    in a straight forward way, as stated in every textbook about
    statistical physics (e.g.\ \cite{nolting2005}).\\
    Because of \(F=U-TS\) where \(U=\avg{H}\) is the internal energy and
    \(S\) entropy, one can guess, that for low \(T\) the internal energy
    will be low, and for big \(T\) the entropy will be high, to minimize
    \(F\). The Hamiltonian of the Ising model connects \(U\) and \(S\)
    by assigning low energies to states with high order. So \(U\) is low
    if all spins are aligned. In this case \(S\) as a measure of
    disorder is also low. Analogical the state of maximum \(U\) is also
    the state of maximum \(S\).
    This preliminary considerations make a phase transition at some
    \(T\), where the influence of the entropy on \(F\) becomes the same
    order of magnitude as the influence of the internal energy on \(F\),
    very plausible.\\
    To determine
    \begin{equation}
        F=-k_{B}T \ln{Z}
    \end{equation}
    one has to know the partition function
    \begin{equation}
        Z = \sum_i p_i = \sum_i e^{-\beta H_i}
    \end{equation}
    where the sum goes over all possible states \(i\) of the system.
    Because every site can have two states, there are \(2^N\) different
    states of the system, which have to be summed up to gain \(Z\).
    Hence for small \(N\) the partition function is computable, but the system
    may show very different properties than in the thermodynamic limit.
    To minimize these finite size effects, it is desirable to examine
    systems with large \(N\). But \(2^N\) is a very rapidly increasing
    number, so there are next to infinity many different states, such
    that it is unfeasible to calculate the energy for each state, except
    for cases, where it is possible to solve it analytically.\\
    If not, one can get estimates of the observables for big \(N\) using
    Monte Carlo simulations, which are introduced in the next chapter.\\
    %~ Once one has obtained values for e.g.\ the magnetization \(m\), one
    %~ can easily calculate the susceptibility
    %~ \begin{equation}
        %~ \chi = \frac{N}{T}\brac{\avg{m^2}-\avg{m}^2}
    %~ \end{equation}
    % Bisschen erklæren, dass man aus m auch chi machen kann und auf literatur verweisen

\subsection{Monte Carlo Simulations}
\label{ssec:montecarlo}
    The idea behind Monte Carlo simulations is to take random samples of
    the observable, which should be measured, and estimate the mean of the observable from
    this samples. In example the Monte Carlo integration chooses random
    points within the bounding box of the function and estimates the integral
    as the fraction of points below the function and points above multiplied
    by the area of the bounding box \cite{Katzgraber2011}.\\
    Analogical the technique can be applied to statistical ensembles,
    but instead of sampling points on an area, one samples states of the
    system.\\
    In statistical physics the expected value of an observable \(O\)
    is calculated by
    \begin{equation}
        \avg{O} = \frac{1}{Z} \sum_i p_i O_i
    \end{equation}
    It is however possible, that there are few states that contribute
    massively more than others. Think of a canonical system at low \(T\),
    where \(p_i\) with low \(H_i\) contribute much more than high \(H_i\).
    But if one samples the \(2^N\) states evenly, it is probable to miss
    them, which possibly introduces big errors to the estimate of \(\avg{O}\).
    It is therefore desirable to sample only the states with high
    contributions to the sum.
    As noted before the system under scrutiny is canonical and therefore
    \(p_i\) is known. The states are distributed according to the Boltzman
    distribution. Hence \emph{Importance Sampling} can be utilized.\\
    Instead of sampling uniformly distributed random states, one samples
    states, which energies are distributed according to a Boltzman distribution.
    In fact this reduces the estimator to
    \begin{equation}
        O_M = \frac{1}{M} \sum_{i=0}^M O_i
    \end{equation}
    as shown in \cite{NewmanBarkema1999}. This is a very convenient form.\\
    But it is difficult to create a random state of a physical system
    e.g.\ the Ising system according to a given distribution. The simple
    method of creating uniformly distributed random states and reject
    them with probability \(p_i^{-1}\) depending on their energy is not
    efficient, because many generated states will be discarded and the
    computing time to generate them and caluculating their energy will
    be wasted.
    Hence one uses \emph{Markov Chains} to generate new states \(\nu\)
    from old ones \(\mu\). It is important that the transition probabilities
    \(A(\mu \to \nu)\) obey \emph{Detailed Balance} and \emph{Ergodicity}.
    \emph{Detailed Balance} means that the probability to leave a state is
    the same as the probability to enter the state in equilibrium
    \(p_\mu A(\mu \to \nu) = p_\nu A(\nu \to \mu)\) with \(p_\mu\) the
    probability to be in state \(\mu\). This ensures that the system can
    equilibrate and that the states are distributed according to the
    desired distribution in equilibrium \cite{NewmanBarkema1999}.
    And \emph{Ergodicity} requires that every possible state is reachable
    from every other state in finite time. \cite{NewmanBarkema1999} \cite{Katzgraber2011}
    Otherwise the samples might not be representative for the whole system.\\
    In this thesis three algorithms, which fulfill all requirements,
    were used. They will be shortly described in the following subsections.

@@

        progressing simulation, as pictured in fig. \ref{fig:equiandauto}\subref{sfig:equiandauto:autoM}.

@@

        the equilibrium is reached after approximately \(N=50\) sweeps for

>>>>

\subsection{Thermodynamic Theory}
\label{ssec:theory}
    The disordered Ising model will be examined as a \emph{canonical system} in
    \emph{equilibrium}. A canonical system can exchange energy with a
    heat bath, thus has a constant temperature -- the temperature of the
    heat bath.
    Equilibrium is defined as a steady state, where
    the observables are only fluctuating but not changing in any
    particular direction. For example thermal equilibrium denotes the
    condition, that the system under scrutiny has reached the temperature
    of the heat bath. This is the case, once no energy is exchanged
    between them, thus the energy of the simulated system reaches a
    steady state.
    In a canonical ensemble the probability \(p_i\) of a state
    \(i\) is distributed according to a Boltzman distribution
    \begin{equation}
        p_i \propto e^{-\beta H_i}
    \end{equation}
    \begin{equation}
        \beta = \frac{1}{k_B T}
    \end{equation}
    Further the free energy \(F\) of a canonical ensemble is minimized
    in equilibrium and all observables can be derived from \(F\)
    in a straight forward way, as stated in every textbook about
    statistical physics (e.g.\ \cite{nolting2005}).\\
    Because of \(F=U-TS\) where \(U=\avg{H}\) is the internal energy and
    \(S\) entropy, one can guess, that for low \(T\) the internal energy
    will be low, and for big \(T\) the entropy will be high, to minimize
    \(F\). The Hamiltonian of the Ising model connects \(U\) and \(S\)
    by assigning low energies to states with high order. So \(U\) is low
    if all spins are aligned. In this case \(S\) as a measure of
    disorder is also low. Analogical the state of maximum \(U\) is also
    the state of maximum \(S\).
    This preliminary considerations make a phase transition at some
    \(T\), where the influence of the entropy on \(F\) becomes the same
    order of magnitude as the influence of the internal energy on \(F\),
    very plausible.\\
    To determine
    \begin{equation}
        F=-k_{B}T \ln{Z}
    \end{equation}
    or any expected value of an observable
    \begin{equation}
        \avg{O} = \frac{1}{Z} \sum_i O_i e^{-\beta H_i}
    \end{equation}
    one has to know the partition function
    \begin{equation}
        Z = \sum_i p_i = \sum_i e^{-\beta H_i}
    \end{equation}
    where the sum goes over all possible states \(i\) of the system.
    Because every site can have two states, there are \(2^N\) different
    states of the system, which have to be summed up to gain \(Z\).
    Hence for small \(N\) the partition function is computable, but the system
    may show very different properties than in the thermodynamic limit.
    To minimize these finite size effects, it is desirable to examine
    systems with large \(N\). But \(2^N\) is a very rapidly increasing
    number, so there are next to infinity many different states, such
    that it is unfeasible to calculate the energy for each state, except
    for cases, where it is possible to solve it analytically.\\
    If not, one can get estimates of the observables for big \(N\) using
    Monte Carlo simulations, which are introduced in the next chapter.\\

    The observables which are measured in this thesis, are the mean
    magnetization per spin
    \begin{equation}
        m = \frac{1}{N} \sum_i s_i
    \end{equation}
    and the mean energy per spin
    \begin{equation}
        E = \frac{1}{N} \avg{H}
    \end{equation}
    As mentioned before, properties near the phase transition and the
    critical temperature \(T_c\), where the phase transition occurs, will
    be examined. The Ising system in two dimensions shows a second order
    phase transition, hence \(m\) and \(E\) are continous, but show at
    \(T_c\) an infinitly sharp slope i.e.\ the first derivative diverges.
    From statistical physics (compare \cite{nolting2005}) is known, that
    these derivatives can be expressed by fluctuations e.g.\ the specific
    heat can be expressed as
    \begin{equation}
        c = \frac{\partial \avg{H}}{\partial T} = k_B \beta^2 \avg{\brac{H-\avg{H}}^2}
    \end{equation}
    The specific heat is a measure for how much energy is needed to change
    the temperature of the system.
    Analogical the susceptibility
    \begin{equation}
        \chi = N \beta \avg{\brac{m-\avg{m}}^2}
    \end{equation}
    is a measure for how strong an outer magnetic field changes the
    magnetization of the system.
    Beside these observables from classical physics, a fifth observable
    the binder cumulant \cite{Binder1981}
    \begin{equation}
        g = \frac{3}{2}\brac{1-\frac{\avg{m^4}}{3\avg{m^2}^2}}
        \label{eq:binder}
    \end{equation}
    is considered.
    This is a dimensionless value, which can be used to determine the
    critical point. These five observables will be sufficient to analyse
    the phase transition in the scope of this bachelor thesis. All of them
    can be easily computed when \(m\) and \(E\) are measured. The next
    chapter will show, how to get estimates for them through Monte Carlo
    simulations.

\subsection{Monte Carlo Simulations}
\label{ssec:montecarlo}
    The idea behind Monte Carlo simulations is to take random samples of
    the observable, which should be measured, and estimate the mean of the observable from
    this samples. As an example the Monte Carlo integration chooses random
    points within the bounding box of the function and estimates the integral
    as the fraction of points below the function and points above multiplied
    by the area of the bounding box \cite{Katzgraber2011}.\\
    Analogical the technique can be applied to statistical ensembles,
    but instead of sampling points on an area, one samples states of the
    system.\\
    In statistical physics the expected value of an observable \(O\)
    is -- as also noted above -- calculated by
    \begin{equation}
        \avg{O} = \frac{1}{Z} \sum_i p_i O_i
    \end{equation}
    It is however possible, that there are few states that contribute
    massively more than others. Think of a canonical system at low \(T\),
    where \(p_i\) with low \(H_i\) contribute much more than high \(H_i\).
    But if one samples the \(2^N\) states evenly, it is probable to miss
    them, which possibly introduces big errors to the estimate of \(\avg{O}\).
    It is therefore desirable to sample only the states with high
    contributions to the sum.
    As noted before the system under scrutiny is canonical and therefore
    \(p_i\) is known. The states are distributed according to the Boltzman
    distribution. Hence \emph{Importance Sampling} can be utilized.\\
    Instead of sampling uniformly distributed random states, one samples
    states, which energies are distributed according to a Boltzman distribution.
    In fact this reduces the estimator to
    \begin{equation}
        O_M = \frac{1}{M} \sum_{i=0}^M O_i
    \end{equation}
    as shown in \cite{NewmanBarkema1999}. This is a very convenient form.\\
    But it is difficult to create a random state of a physical system
    e.g.\ the Ising system according to a given distribution. The simple
    method of creating uniformly distributed random states and reject
    them with probability \(p_i^{-1}\) depending on their energy is not
    efficient, because many generated states will be discarded and the
    computing time to generate them and caluculating their energy will
    be wasted.
    Hence one uses \emph{Markov Chains} to generate new states \(\nu\)
    from old ones \(\mu\). It is important that the transition probabilities
    \(A(\mu \to \nu)\) obey \emph{Detailed Balance} and \emph{Ergodicity}.
    \emph{Detailed Balance} means that the probability to leave a state is
    the same as the probability to enter the state in equilibrium
    \(p_\mu A(\mu \to \nu) = p_\nu A(\nu \to \mu)\) with \(p_\mu\) the
    probability to be in state \(\mu\). This ensures that the system can
    equilibrate and that the states are distributed according to the
    desired distribution in equilibrium \cite{NewmanBarkema1999}.
    And \emph{Ergodicity} requires that every possible state is reachable
    from every other state in finite time. \cite{NewmanBarkema1999} \cite{Katzgraber2011}
    Otherwise the samples might not be representative for the whole system.\\
    In this thesis three algorithms, which fulfill all requirements,
    were used. They will be shortly described in the following subsections.

@@

        progressing simulation, as pictured in fig. \ref{fig:equiandauto}\subref{sfig:equiandauto:equiE}.

@@

        the equilibrium is reached after approximately \(N_{s} \approx 50\) sweeps for
